---
title: "Data Cleaning for per-pathogen virulence project"
author: "Judith Bouman"
date: "2023-01-09"
output: pdf_document
---
```{r setup, include=FALSE}
# Load neccasery libraries

library(lmerTest)
library(psych)
library(ggplot2)
library(grDevices)
library(lme4)
library(optimx)
library(scatterplot3d) 
library(qpcR)
library(chngpt)
library(bbmle)
library(lme4)
library(matlib)
library(numDeriv)

# put settings to enghlish, to get the correct dates afterwards.
Sys.setlocale("LC_TIME", "English")

# Set directory
setwd("/Users/jb22m516/Documents/GitHub/ppv")

```

## Read in data

This document contains all code to reproduce the analysis described in the manuscript.  
The first step is to read all data. 

```{r read in data, echo = F}
# Read in data files
VL_data = read.csv('vl.csv',header = TRUE, blank.lines.skip = TRUE) # viral load data
CD4_data = read.csv('cd4prof_NEW.csv',header = TRUE, blank.lines.skip = TRUE) # CD4 data --> do I even need these?
data = read.csv('data_perparacite_copy.csv',header = TRUE, blank.lines.skip = TRUE) # original datafile with data combined
arv = read.csv('arv.csv',header = TRUE, blank.lines.skip = TRUE) # data on treatments used

attach(data)

# treat data such that it can be used

# Remove d from time measure and convert to integer
data$Time..Days. = as.integer(gsub( "d", "", as.character(data$Time..Days.)))

# Remove empty rows
data = data[!is.na(data$Patient.ID),]

# fix names of columns
names(VL_data) <- c('PTID', names(VL_data)[2:11])
names(CD4_data) <- c('PTID', names(CD4_data)[2:5]) 
names(data) <- c('Patient.ID', names(data)[2:27])

# Determine which IDs to use in the analysis (firs instance use only from the original analysis) CHANGE LATER
IDs_VL = sort(unique(VL_data$PTID))
IDs_CD4 = sort(unique(CD4_data$PTID)) 
IDs_data = sort(unique(data$Patient.ID))
IDs = unique(c(IDs_CD4,IDs_data,IDs_VL))
```

## Methods

# Inclusion criteria 

The next part of the code creates a list of all viral load measures and CD4 counts per individual that were collected after the primary infection. 
The duration of the primary infection is set to be 90 days. 
The code also creates a list of the subtypes, ages and country of origin for all participants. 

Moreover, it checks for each participant whether the data fulfills the requirements of inclusion: 
1) at least 4 CD4 measurements, 
2) at least 3 viral load measurments, 
3) measurements of CD4 count are at least 180 days apart, 
4) the subtype of the individual is known. 

```{r filter data, echo=FALSE}
#######################################################################
################# Make per patient list of all VL and CD4 #############
#######################################################################

# initialize vectors
per_patient_VL <- rep( list(0), length(IDs))        # all VL data per patient
per_patient_CD4 <- rep( list(0), length(IDs))       # all CD4 data per patient

subtypes =  vector(mode = "character", length(IDs)) # patient subtype
country = vector(mode = "character", length(IDs))   # Country
age = vector(mode = "numeric", length(IDs))         # age of patient
ART1DT = vector(mode = "numeric", length(IDs))      # start of ART treatment if appicable
suitable = rep(FALSE, length(IDs))                  # vector indicating if enough data is available for each patient

# Do you want to have a general primary infection period (1) or a subtype specific (2)? In the final manuscript we decided to not report on option 2 due to data restrictions. 
primary = 1

i = 1

for (j in IDs){

  tot = data[data$Patient.ID==j,]

  per_patient_VL[[i]] = VL_data[VL_data$PTID==j,c(2,3,4)]
  per_patient_CD4[[i]] = CD4_data[CD4_data$PTID==j,c(2,3,4)]

  #print(per_patient_VL[[i]])
  #print(per_patient_CD4[[i]])

  # Remove CD4 counts without a date
  per_patient_CD4[[i]] = per_patient_CD4[[i]][!per_patient_CD4[[i]]$VISDATE=="",]

  per_patient_VL[[i]]$GCLPLAMA[per_patient_VL[[i]]$GCLPLAMA==0] <- 50 # set to detection limit

  # remove NA values for counts
  per_patient_VL[[i]] = per_patient_VL[[i]][!is.na(per_patient_VL[[i]]$GCLPLAMA),]
  per_patient_CD4[[i]] = per_patient_CD4[[i]][!is.na(per_patient_CD4[[i]]$CD4),]

  # make number of days instead of visit number using the dates
  date_VL = as.Date(per_patient_VL[[i]][,2],format='%d.%m.%y')
  date_CD4 = as.Date(per_patient_CD4[[i]][,2],format='%d.%b.%y')

  #determine date of first measurement
  first_date = min(date_CD4, date_VL)

  # add number of days to data
  days_VL = as.numeric(date_VL-first_date)
  per_patient_VL[[i]]$Days <- days_VL
  days_CD4 = as.numeric(date_CD4-first_date)
  per_patient_CD4[[i]]$Days <- days_CD4

  # Determine age, country and subtype
  age[i] = CD4_data[CD4_data$PTID==j,5][1]
  country[i] = as.character(tot$Source[1])
  subtypes[i] = as.character(tot$Patient.Subtype[1]) # can this be improved?

  # in case first subtype has no value, try second one
  if (length(subtypes[i])>0){
    if (!is.na(subtypes[i])){
      if (subtypes[i]==""){
        subtypes[i] = as.character(tot$Patient.Subtype[2])
      }
    }
  }

  # get initial date of treatment initiation, when symptoms where observed
  treatment = floor(arv[arv$PTID==j,]$ART1VIS)

  if (length(treatment)>0){
    ART1DT[i] <- as.numeric(treatment[1])
  } else {
    ART1DT[i] <- NA
  }

  low_CD4_tr = 100 # to make sure it has a value even if it did not happen

  # Also remove patients that score two concecutive times lower than 200
  low.counts = per_patient_CD4[[i]][per_patient_CD4[[i]]$CD4<200,]

  if (dim(low.counts)[1]>1){
    for (k in seq(1,dim(low.counts)[1]-1,1)){
      if ((as.integer(row.names(low.counts))[k]-as.integer(row.names(low.counts))[k+1])==-1){
        low_CD4_tr = low.counts$VISIT[k+1]
        per_patient_CD4[[i]] = per_patient_CD4[[i]][per_patient_CD4[[i]]$VISIT<=low_CD4_tr,]
      }
    }
  }

  # Remove data after smallest visit number of both treatment start possibilities
  min_treatment <- min(ART1DT[i],low_CD4_tr,na.rm = TRUE)

  per_patient_CD4[[i]]=per_patient_CD4[[i]][per_patient_CD4[[i]]$VISIT<=min_treatment,]
  per_patient_VL[[i]]=per_patient_VL[[i]][per_patient_VL[[i]]$VISIT<=min_treatment,]

  # Filter VL data
  # remove points obtained during primary infection (time differs per subtype)
  if (primary == 1){
    t_primary = 90
  } else if (primary == 2){
    if (!is.na(subtypes[i])){
     if (!subtypes[i]==""){
        if (subtypes[i]=='A'){
          t_primary = 139
        } else if (subtypes[i]=='C'){
         t_primary = 183
        } else if (subtypes[i]=='D'){
          t_primary = 100
        }
      }
    }
  }

  per_patient_VL[[i]]=per_patient_VL[[i]][days_VL>t_primary,]
  per_patient_CD4[[i]]=per_patient_CD4[[i]][days_CD4>t_primary,]

  # Determine if participant has enough data to use for analysis
  per_patient_VL[[i]] <- per_patient_VL[[i]][!is.na(per_patient_VL[[i]]$VISIT),]
  per_patient_CD4[[i]] <- per_patient_CD4[[i]][!is.na(per_patient_CD4[[i]]$VISIT),]

  # are there still enough measurements?
  if (dim(per_patient_CD4[[i]])[1]>0 && dim(per_patient_VL[[i]])[1]>0){
    # Is first CD4 measurement within first year?
    # are there at least 3 CD4 measurements?
    if (length(per_patient_CD4[[i]]$CD4)>3){
      # are the at least 180 days apart?
      if ((per_patient_CD4[[i]]$Days[length(per_patient_CD4[[i]]$Days)]-per_patient_CD4[[i]]$Days[1])>180){
        # are there at least 2 VL measures?
        if (length(per_patient_VL[[i]])>2){
          # is subype known?
          if (subtypes[i] %in% c('A','C','D')){
            suitable[i] = TRUE
          }
        }
      }
    }
  }

  i = i+1
}


#### Select only suitable participants
IDs_suit = IDs[suitable]

per_patient_CD4_suit = per_patient_CD4[suitable]
per_patient_VL_suit = per_patient_VL[suitable]

subtypes_suit = subtypes[suitable]
country_suit = country[suitable]

age_suit = age[suitable]
```

I then create a table that summarizes the data of the included the participants. 

```{r summarize data included in the study, echo=FALSE}

tot_par = length(IDs_suit)
num_Uganda = sum(country_suit=='Uganda',na.rm = TRUE)
num_Zimbabwe = sum(country_suit=='Zimbabwe', na.rm = TRUE)
num_a = sum(subtypes_suit=='A', na.rm = TRUE)
num_c = sum(subtypes_suit=='C', na.rm = TRUE)
num_d = sum(subtypes_suit=='D', na.rm = TRUE)
sub_Uganda_A = sum(subtypes_suit[country_suit=='Uganda']=='A',na.rm=TRUE)
sub_Uganda_C = sum(subtypes_suit[country_suit=='Uganda']=='C',na.rm=TRUE)
sub_Uganda_D = sum(subtypes_suit[country_suit=='Uganda']=='D',na.rm=TRUE)
sub_Zimbabwe_A = sum(subtypes_suit[country_suit=='Zimbabwe']=='A',na.rm=TRUE)
sub_Zimbabwe_C = sum(subtypes_suit[country_suit=='Zimbabwe']=='C',na.rm=TRUE)
sub_Zimbabwe_D = sum(subtypes_suit[country_suit=='Zimbabwe']=='D',na.rm=TRUE)

num_CD4 = rep(0, length(IDs_suit))
num_spVL = rep(0, length(IDs_suit))

for (i in seq(1,length(IDs_suit),1)){
  num_CD4[i] = dim(per_patient_CD4_suit[[i]])[1]
  num_spVL[i] = dim(per_patient_VL_suit[[i]])[1]
}

CD4_Uganda_A = median(num_CD4[country_suit=='Uganda'&subtypes_suit=='A'],na.rm=TRUE)
CD4_Uganda_C = median(num_CD4[country_suit=='Uganda'&subtypes_suit=='C'],na.rm=TRUE)
CD4_Uganda_D = median(num_CD4[country_suit=='Uganda'&subtypes_suit=='D'],na.rm=TRUE)
CD4_Zimbabwe_A = median(num_CD4[country_suit=='Zimbabwe'&subtypes_suit=='A'],na.rm=TRUE)
CD4_Zimbabwe_C = median(num_CD4[country_suit=='Zimbabwe'&subtypes_suit=='C'],na.rm=TRUE)
CD4_Zimbabwe_D = median(num_CD4[country_suit=='Zimbabwe'&subtypes_suit=='D'],na.rm=TRUE)

VL_Uganda_A = median(num_spVL[country_suit=='Uganda'&subtypes_suit=='A'],na.rm=TRUE)
VL_Uganda_C = median(num_spVL[country_suit=='Uganda'&subtypes_suit=='C'],na.rm=TRUE)
VL_Uganda_D = median(num_spVL[country_suit=='Uganda'&subtypes_suit=='D'],na.rm=TRUE)
VL_Zimbabwe_A = median(num_spVL[country_suit=='Zimbabwe'&subtypes_suit=='A'],na.rm=TRUE)
VL_Zimbabwe_C = median(num_spVL[country_suit=='Zimbabwe'&subtypes_suit=='C'],na.rm=TRUE)
VL_Zimbabwe_D = median(num_spVL[country_suit=='Zimbabwe'&subtypes_suit=='D'],na.rm=TRUE)

age_Uganda_A = median(age_suit[country_suit=='Uganda'&subtypes_suit=='A'],na.rm=TRUE)
age_Uganda_C = median(age_suit[country_suit=='Uganda'&subtypes_suit=='C'],na.rm=TRUE)
age_Uganda_D = median(age_suit[country_suit=='Uganda'&subtypes_suit=='D'],na.rm=TRUE)
age_Zimbabwe_A = median(age_suit[country_suit=='Zimbabwe'&subtypes_suit=='A'],na.rm=TRUE)
age_Zimbabwe_C = median(age_suit[country_suit=='Zimbabwe'&subtypes_suit=='C'],na.rm=TRUE)
age_Zimbabwe_D = median(age_suit[country_suit=='Zimbabwe'&subtypes_suit=='D'],na.rm=TRUE)

rows = c('total number of patients included', 'number of patients from Zimbabwe',
         'number of patients from Uganda', 'number of patients with type A', 'number of patients with type C',
         'number of patients with type D', 'number of patients in Zimbabwe with type A', 'number of patients in Zimbabwe with type C',
         'number of patients in Zimbabwe with type D',  'number of patients in Uganda with type A',
         'number of patients in Uganda with type C',  'number of patients in Uganda with type D')

values = c(num_Uganda+num_Zimbabwe,num_Zimbabwe, num_Uganda, num_a+1, num_c,
           num_d, sub_Zimbabwe_A, sub_Zimbabwe_C, sub_Zimbabwe_D, sub_Uganda_A, sub_Uganda_C, sub_Uganda_D)

summary_data = matrix(values,nrow = length(rows))
row.names(summary_data)<-rows

summary_data

```

# Calculating setpoint viral load and decline of CD4 T cells 

Based on the measurements per individual we calculate the setpoint viral load by taking the geometric mean of all included viral load measurements. 
Also, I calculate the CD4 change by taking the slope of a linear fit through the data of each individual separately. To check that a linear fit through the data is the most suitable, I compare this fit with a:
two-phase linear, 
exponential,
and power-law fit. 

```{r calculate setpoint VL and CD4 decline, echo=FALSE}

setpoint_VL = vector(mode = "numeric", length(IDs_suit)) # setpoint VL measure per patient
CD4_decline = vector(mode = "numeric", length(IDs_suit)) # CD4 decline measure per patient

num_measurements = vector(mode = "numeric", length(IDs_suit))
CD4_int = vector(mode = "numeric", length(IDs_suit))
CD4_first = vector(mode = "numeric", length(IDs_suit))
CD4_first_time = vector(mode = "numeric", length(IDs_suit))
span_CD4 = vector(mode = "numeric", length(IDs_suit))
meas_CD4 = vector(mode = "numeric", length(IDs_suit))

for (i in seq(1,length(IDs_suit),1)){
  setpoint_VL[i] = geometric.mean(per_patient_VL_suit[[i]]$GCLPLAMA, na.rm = TRUE)
  fit = lm(per_patient_CD4_suit[[i]]$CD4 ~ per_patient_CD4_suit[[i]]$Days)
  CD4_decline[i] = fit$coefficients[[2]]
  CD4_int[i] = fit$coefficients[[1]]
  CD4_first[i]=per_patient_CD4_suit[[i]]$CD4[1]
  CD4_first_time[i]=per_patient_CD4_suit[[i]]$Days[1]
  span_CD4[i] = per_patient_CD4_suit[[i]]$Days[length(per_patient_CD4_suit[[i]]$Days)]-per_patient_CD4_suit[[i]]$Days[1]
  meas_CD4[i] = length(per_patient_CD4_suit[[i]]$Days)
}

```

Additionally, we do a mixed linear effect model to calculate the CD4 decline of each individual. 
We also use that model to calculate the intercept of each individual. 

```{r mixed effect model, echo=FALSE}
library('lattice')
## Estimate CD4 declines using lmer
name_ID <- rep('', length(IDs_suit))

for (i in seq(1,length(IDs_suit),1)){
  if (i<10){
    name_ID[i] <- paste('ID','00',i, sep='')
  } else if (i<100){
    name_ID[i] <- paste('ID','0',i, sep='')
  } else if (i<1000){
    name_ID[i] <- paste('ID',i, sep='')
  }
}

CD4_all = c()
Days_all = c()
IDs_tot = c()
subtypes_all = c()
ages_all = c()
country_all = c()
spVL_all = c()
init_all = c()
VL_all = c()

for (i in seq(1,length(IDs_suit),1)){
  test = per_patient_CD4_suit[[i]]
  CD4_all = c(CD4_all,test$CD4)
  Days_all = c(Days_all,test$Days)
  IDs_tot <- c(IDs_tot,rep(name_ID[i],length(test$CD4)))
  subtypes_all <- c(subtypes_all,rep(subtypes_suit[i],length(test$CD4)))
  ages_all <- c(ages_all,rep(age_suit[i],length(test$CD4)))
  country_all <- c(country_all,rep(country_suit[i],length(test$CD4)))
  spVL_all <- c(spVL_all,rep(setpoint_VL[i],length(test$CD4)))
  init_all <- c(init_all, rep(CD4_first[i], length(test$CD4)))
  VL_all <- c(VL_all, per_patient_VL_suit[[i]]$GCLPLAMA )
}

Days_all_rescale <- scale(Days_all, center=FALSE)

# basic linear model
lmm_fit_CD4 <- lmer(CD4_all~Days_all_rescale + (Days_all_rescale|IDs_tot), REML=FALSE)
#summary(lmm_fit_CD4)
#coef(lmm_fit_CD4)
R_CD4 <-  1 - ( sum( (CD4_all-predict(lmm_fit_CD4))^2 ) )/(sum( (CD4_all-mean(CD4_all))^2 ))
R_CD4

```

We test, by comparing 4 different models (linear, exponential, power-law, 2 phase linear fit) if a linear fit is indeed best to fit the CD4 T cell data. 

```{r decide on individual model of decline, echo=FALSE}
## compare fit to different kinds of model
ni <- rep(0, length(IDs_suit))

RSS_linear <- rep(0, length(IDs_suit))
RSS_exponential <- rep(0, length(IDs_suit))
RSS_power <- rep(0, length(IDs_suit))
RSS_2phase <- rep(0, length(IDs_suit))

for (i in seq(1,length(IDs_suit),1)){
  test = per_patient_CD4_suit[[i]]
  
  # linear model 
  linear_test <- lm(test$CD4 ~ test$Days )
  RSS_linear[i] <- RSS(linear_test)
  ni[i] <- dim(test)[1]

  # two-phase linear model
  # only perform if number of points is larger than ... otherwise use single linear fit
  # check which break point is the most likely (use each data point as possibility)
  intermediate.RSS <- rep(NA, length(2:(dim(test)[1]-1)))
  
  for (j in 2:(dim(test)[1]-1)){
    # Fit first points 
    fit1 <- lm( CD4~Days, data=test[1:j,] )
    # time until which we use fit 1: 
    t.dif <- (test$Days[j]+ test$Days[j+1])/2
    int.dif <- fit1$coefficients[1]+ t.dif*fit1$coefficients[2]
    # fit second linear model to rest of data
    data.2 <- test[(j+1):dim(test)[1],]
    Days.2 <- data.2$Days - t.dif
    CD4.2 <- data.2$CD4 - int.dif
    fit2 <- lm( CD4.2 ~ Days.2 + 0 )
    
    intermediate.RSS[j-1] <- RSS(fit1) + RSS(fit2)
  }
  
  RSS_2phase[i] <- min(intermediate.RSS)
  
  # power law model
  power_test <- lm(log10(test$CD4) ~ log10(test$Days))
  RSS_power[i] <- sum((test$CD4 - 10^power_test$fitted.values)^2)
  
  # exponential model 
  exponential_test <- lm(log10(test$CD4) ~ test$Days )
  RSS_exponential[i] <- sum((test$CD4 - 10^exponential_test$fitted.values)^2)
}

#number of parameters per model + 1
k_linear = 2 + 1
k_power = 2 + 1
k_exponential = 2 + 1
k_2phase = 2+1+1+1 #two times a linear fit, once with intercept + the point where the fit breaks  

# Calculate AICc value for each model over all individual data sets
AIC_linear <- sum(ni* log(RSS_linear/ni) ) + 2*k_linear*length(IDs_suit)*(sum(ni))/(sum(ni)-k_linear*length(IDs_suit)-length(IDs_suit)) #46140.72
AIC_power <- sum(ni* log(RSS_power/ni) ) + 2*k_power*length(IDs_suit)*(sum(ni))/(sum(ni)-k_power*length(IDs_suit)-length(IDs_suit)) #46902.91
AIC_exponential <- sum(ni* log(RSS_exponential/ni) ) + 2*k_exponential*length(IDs_suit)*(sum(ni))/(sum(ni)-k_exponential*length(IDs_suit)-length(IDs_suit)) #46413.87
AIC_2phase <- sum(ni* log(RSS_2phase/ni) ) + 2*k_2phase*length(IDs_suit)*(sum(ni))/(sum(ni)-k_2phase*length(IDs_suit)-length(IDs_suit)) #46712.27

############################################################
```

Now, I create a file that contains all data of Figure 2, such that we can share this data. 

```{r publish data, echo=FALSE}

# Calculate rate of disease progression, the CD4 level 
general_int = coef(lmm_fit_CD4)$IDs_tot[,1]
general_int_raw = coef(lmm_fit_CD4)$IDs_tot[,1]
general_slope = coef(lmm_fit_CD4)$IDs_tot[,2]/(max(Days_all)/max(Days_all_rescale)) * 365
general_slope_raw = coef(lmm_fit_CD4)$IDs_tot[,2]
CD4_int <- rep(0,length(IDs_suit))

# correct for eliminating measures of the first 90 days of the measurements
CD4_int[subtypes_suit=='A'] <- general_int[subtypes_suit=='A']  + 90/((max(Days_all)/max(Days_all_rescale)))*general_slope[subtypes_suit=='A']
CD4_int[subtypes_suit=='C'] <- general_int[subtypes_suit=='C']  + 90/((max(Days_all)/max(Days_all_rescale)))*general_slope[subtypes_suit=='C']
CD4_int[subtypes_suit=='D'] <- general_int[subtypes_suit=='D']  + 90/((max(Days_all)/max(Days_all_rescale)))*general_slope[subtypes_suit=='D']

data_open = cbind( subtype = subtypes_suit, 
                   countryOO = country_suit, 
                   age = age_suit, 
                   CD4_0 = CD4_int, 
                   CD4_raw = coef(lmm_fit_CD4)$IDs_tot[,1],
                   DCD4_perYear = general_slope, 
                   DCD4_raw = general_slope_raw, 
                   spVL = setpoint_VL)

write.csv2( data_open , file = "/Users/jb22m516/OneDrive - Universitaet Bern/Project_B/DataSharing_PPV/data_open.csv" )

```