---
title: "Main Analysis of manuscript -- Per-pathogen Virulence of HIV-1 subtypes A, C and D"
author: "Judith Bouman"
date: "2023-01-09"
output: pdf_document
---

# Setup

Setup R and read open data file.

```{r setup, include=FALSE}
# Load neccasery libraries

library(lmerTest)
library(psych)
library(ggplot2)
library(grDevices)
library(lme4)
library(optimx)
library(scatterplot3d) 
library(qpcR)
library(chngpt)
library(bbmle)
library(lme4)
library(matlib)
library(numDeriv)
library(ggpubr)

# put settings to enghlish, to get the correct dates afterwards.
Sys.setlocale("LC_TIME", "English")

path_WD = "/Users/jb22m516/OneDrive - Universitaet Bern/Project_B/DataSharing_PPV/"

```

```{r read data, echo=FALSE, include = FALSE}

data = read.csv2( file = paste0(path_WD, "data_open.csv") )

data$CD4_0 = as.numeric(data$CD4_0)
data$CD4_raw = as.numeric(data$CD4_raw)
data$DCD4_perYear = as.numeric(data$DCD4_perYear)
data$DCD4_raw = as.numeric(data$DCD4_raw)
data$spVL = as.numeric(data$spVL)

subtypes_suit = data$subtype
country = data$countryOO

```

# Analyses

Calculate disease progression from the decline and initial values after primary infection.

```{r calculate rate of disease progression, echo=FALSE, include = FALSE}

gamma <- rep(0,dim(data)[1])
gamma[data$subtype=='A'] <- 1/(  (( (200-data$CD4_0[data$subtype=='A']) / data$DCD4_perYear[data$subtype=='A']) ) + 90/365 )    # and scale back to years instead of rescaled days
gamma[data$subtype=='C'] <- 1/(  (( (200-data$CD4_0[data$subtype=='C']) / data$DCD4_perYear[data$subtype=='C']) ) + 90/365 )  
gamma[data$subtype=='D'] <- 1/(  (( (200-data$CD4_0[data$subtype=='D']) / data$DCD4_perYear[data$subtype=='D']) ) + 90/365 )  

cor.test( data$CD4_raw, data$DCD4_raw )

```

## Create Figure 2

Compare the CD4 cell decline (calculated with the mixed linear model), the CD4 level at the start of the chronic infection (calculated with the mixed linear model), the setpoint viral load and the disease progression between the subtypes. 

```{r generate boxplots, echo=FALSE, fig.height=8, fig.width=8, include = FALSE}

# Make boxplots of different observations
wilcox.test(data$DCD4_perYear[subtypes_suit=='A'], data$DCD4_perYear[subtypes_suit=='C']) # p-value = 0.1758
wilcox.test(data$DCD4_perYear[subtypes_suit=='A'], data$DCD4_perYear[subtypes_suit=='D']) # p-value = 0.3867
wilcox.test(data$DCD4_perYear[subtypes_suit=='C'], data$DCD4_perYear[subtypes_suit=='D']) # p-value = 0.03737

wilcox.test(data$CD4_0[subtypes_suit=='A'], data$CD4_0[subtypes_suit=='C'])
wilcox.test(data$CD4_0[subtypes_suit=='D'], data$CD4_0[subtypes_suit=='C']) 
wilcox.test(data$CD4_0[subtypes_suit=='A'], data$CD4_0[subtypes_suit=='D']) 

wilcox.test(data$CD4_0[country=='Uganda'], data$CD4_0[country=='Zimbabwe']) 

wilcox.test(gamma[subtypes_suit=='A'], gamma[subtypes_suit=='C']) #  
wilcox.test(gamma[subtypes_suit=='A'], gamma[subtypes_suit=='D']) #  
wilcox.test(gamma[subtypes_suit=='C'], gamma[subtypes_suit=='D']) # 

wilcox.test(data$spVL[subtypes_suit=='A'], data$spVL[subtypes_suit=='C']) #   p-value = 0.3976
wilcox.test(data$spVL[subtypes_suit=='A'], data$spVL[subtypes_suit=='D']) #   p-value = 0.07222
wilcox.test(data$spVL[subtypes_suit=='C'], data$spVL[subtypes_suit=='D']) #   p-value = 0.0194

# Make figures
data_2A = data.frame(data$DCD4_perYear, subtypes_suit)
fig2_A <- ggplot(data_2A, aes(x=subtypes_suit, y=data$DCD4_perYear)) +
  geom_point(aes(fill=subtypes_suit), alpha=0.5, size=2, shape=21,
             position=position_jitter(width=0.2, height=0)) +
  geom_boxplot(outlier.colour=NA, fill=NA)+
  xlab("HIV-1 subtype") + ylab(expression(paste(Delta, "CD4 ", "(cells ", mu, "l/year)")))  + labs(fill = "") +
  scale_fill_manual(values=c("gold", "#BE583B","#6389A7")) + theme_bw(base_size = 14) +
  theme(legend.position="none")

data_2B = data.frame(data$CD4_0, subtypes_suit)
fig2_B <- ggplot(data_2B, aes(x=subtypes_suit, y=data$CD4_0 )) +
  geom_point(aes(fill=subtypes_suit), alpha=0.5, size=2, shape=21,
             position=position_jitter(width=0.2, height=0)) +
  geom_boxplot(outlier.colour=NA, fill=NA)+
  xlab("HIV-1 subtype")  + ylab(expression(paste("CD4 count at the start of chronic inf. (cells ", mu, "l/year)"))) + labs(fill = "") +
  scale_fill_manual(values=c("gold", "#BE583B","#6389A7")) + theme_bw(base_size = 14) +
  theme(legend.position="none")

data_2C = data.frame(gamma, subtypes_suit)
fig2_C <- ggplot(data_2C, aes(x=subtypes_suit, y=gamma)) +
  geom_point(aes(fill=subtypes_suit), alpha=0.5, size=2, shape=21,
             position=position_jitter(width=0.2, height=0)) +
  geom_boxplot(outlier.colour=NA, fill=NA)+
  xlab("HIV-1 subtype") + ylab('Rate of disease progr. (per year)')  + labs(fill = "") +
  scale_fill_manual(values=c("gold", "#BE583B","#6389A7")) + theme_bw(base_size = 14) +
  theme(legend.position="none")

data_2D = data.frame(data$spVL, subtypes_suit)
fig2_D <- ggplot(data_2D, aes(x=subtypes_suit, y=log10(data$spVL) )) +
  geom_point(aes(fill=subtypes_suit), alpha=0.5, size=2, shape=21,
             position=position_jitter(width=0.2, height=0)) +
  geom_boxplot(outlier.colour=NA, fill=NA)+
  xlab("HIV-1 subtype") + ylab('spVL (log10 (RNA copies/ml))')  + labs(fill = "") +
  scale_fill_manual(values=c("gold", "#BE583B","#6389A7")) + theme_bw(base_size = 14) +
  theme(legend.position="none")

# save figure 2 
pdf(paste0(path_WD, "Figure_2.pdf"), width = 10, height = 12)
ggarrange(fig2_A, fig2_B, fig2_D, fig2_C, labels = c("A", "B", 'C','D'),
                    ncol = 2, nrow = 2)
dev.off()
```

## Create Supplementary Figure 1

Country-wise comparison. 

```{r generate SF1, echo=FALSE, fig.height=8, fig.width=8, include = FALSE}
# save supplementary figure 1 --> 
data_S1 = data.frame(data$CD4_0, country)
suppl_1 <- ggplot(data_S1, aes(x=country, y=data$CD4_0 )) +
  geom_point(aes(fill=country), alpha=0.5, size=2, shape=21,
             position=position_jitter(width=0.2, height=0)) +
  geom_boxplot(outlier.colour=NA, fill=NA)+
  xlab("Country of origin") + ylab(expression(paste("CD4 count at the start of chronic inf. (cells ", mu, "l/year)"))) + labs(fill = "") +
  scale_fill_manual(values=c("orange", "green")) + theme_bw(base_size = 14) +
  theme(legend.position="none")

pdf( paste0(path_WD, "SupplementaryF1.pdf"), width = 5, height = 6)
suppl_1
dev.off()

```

## Results for Table 2

Next, we model the relationship between disease progression (gamma) and the setpoint viral load, to find the per-pathogen virulence. 

```{r analyse relationship between gamma and spVL, echo=FALSE, include = FALSE, include = FALSE}

# Should we consider an exponential fit?
# create a vector with the relative ages: such that the mean age is equal to zero:
age_rel <- data$age - mean(data$age)

c_all <- rep(0, length(data$age))
c_all[subtypes_suit=="C"] <- 1

a_all <- rep(0, length(data$age))
a_all[subtypes_suit=="A"] <- 1

d_all <- rep(0, length(data$age))
d_all[subtypes_suit=="D"] <- 1

data_fitting <- data.frame(gamma, spvl = log10(data$spVL), age_rel,  a_all, c_all, d_all)

# Alternative approach: fit everything in one using the MLE approach, because the nls functions does not produce the correct uncertainty
LL <- function( q=1e-5, a_c = 0.0001, d_a = -0.0001, d_d = 0.00001, c = 2, sd_fit = 0.1){
  -sum(stats::dnorm(0, mean=(gamma - (q*age_rel + a_c + d_a*a_all + d_d*d_all)*(spvl)^(c) ), sd=sd_fit, log=TRUE))
}

fit <- mle2(LL , data=data_fitting  )
summary(fit)
confint(fit)


# Use hessian method to calculate confidence interval for c 
LL_hes <- function(x = c( q=1e-5, a_c = 0.0001, d_a = -0.0001, d_d = 0.00001, c = 2, sd_fit = 0.1), data = data_fitting ){
  -sum(stats::dnorm(0, mean=(data$gamma - (x["q"]*data$age_rel + x["a_c"] + x["d_a"]*data$a_all + x["d_d"]*data$d_all)*(data$spvl)^(x["c"]) ), sd=x["sd_fit"], log=TRUE))
}

# confidence interval for c ( because the profile gives an NA)
hess <- hessian(LL_hes, x = fit@coef)
sd_c <- sqrt(diag(inv(hess))[5])
fit@coef["c"]+1.96*sd_c
fit@coef["c"]-1.96*sd_c

predict_fit <- (coef(fit)[[1]]*data_fitting$age_rel + coef(fit)[[2]] + coef(fit)[[3]]*data_fitting$a_all + 
                  coef(fit)[[4]]*data_fitting$d_all)*(data_fitting$spvl)^(coef(fit)[[5]])

# calculate R squared
R_sqrt = 1 - ( sum( (data_fitting$gamma-predict_fit)^2 ) )/(sum( (data_fitting$gamma-mean(data_fitting$gamma))^2))
R_sqrt 

```

## Suplementary Figure 2

```{r FS2, echo=FALSE, fig.height=6, fig.width=8, include = FALSE}

line = 1
cex = 2
side = 3
adj=-0.05

pdf(paste0(path_WD, "SupplementaryF2.pdf"), width = 14, height = 7)
{par(mfrow=c(1,2), oma=c(1,6,1,1)) 
qqnorm(data_fitting$gamma-predict_fit, pch = 1, frame = FALSE)
mtext("A", side=side, line=line, cex=cex, adj=adj)
hist(data_fitting$gamma-predict_fit, n=50, xlab="Model residuals", main="")
mtext("B", side=side, line=line, cex=cex, adj=adj) }
dev.off()
```

## Figure 3

```{r Figure 3, echo=FALSE, fig.height=6, fig.width=8, include = FALSE}

# make a graph of the results 
pdf(paste0(path_WD, "Figure_3.pdf"), width = 8, height = 6)
x <- seq(0,6, 0.1)
df.A <- data.frame(10^x, rep(0,length(x)), rep('A', length(x)))
colnames(df.A) <- c('spVL','age_rel','subtypes')
df.A$predict_fit <- (coef(fit)[[2]] + coef(fit)[[3]])*(log10(df.A$spVL))^(coef(fit)[[5]])
df.C <- data.frame(10^x, rep(0,length(x)), rep('C', length(x)))
colnames(df.C) <- c('spVL','age_rel','subtypes')
df.C$predict_fit <- (coef(fit)[[2]])*(log10(df.C$spVL))^(coef(fit)[[5]])
df.D <- data.frame(10^x, rep(0,length(x)), rep('D', length(x)))
colnames(df.D) <- c('spVL','age_rel','subtypes')
df.D$predict_fit <- (coef(fit)[[2]] + coef(fit)[[4]])*(log10(df.D$spVL))^(coef(fit)[[5]])

plot(log10(data$spVL)[subtypes_suit=='A'],gamma[subtypes_suit=='A'], ylab = 'Rate of disease progression (1/year)', xlab = 'log10( spVL )',
     ylim = c(min(gamma),max(gamma)), col='gold', pch=19, xlim=c(1.7,5.8),cex.lab=1.5,cex.axis=1.5)
points(log10(data$spVL)[subtypes_suit=='C'],gamma[subtypes_suit=='C'],
       col='#BE583B', pch=19)
points(log10(data$spVL)[subtypes_suit=='D'],gamma[subtypes_suit=='D'],
       col='#6389A7', pch=19)

lines(x, df.A$predict_fit,
      lwd=3,col='gold', cex=1.5)
lines(x, df.C$predict_fit,
      lwd = 3,col='#BE583B')
lines(x, df.D$predict_fit,
      lwd=3,col='#6389A7')
legend('topleft', c('Subtype A', 'Subtype C', 'Subtype D'), lty = c(1,1,1), col=c("gold", "#BE583B","#6389A7"), lwd=3, bty='n')

lines(seq(min(log10(data$spVL))-0.5,max(log10(data$spVL))+0.5,0.01), rep(1/10, length(seq(min(log10(data$spVL))-0.5,max(log10(data$spVL))+0.5,0.01) ) ),
      lwd=1 , col = 'grey' )
text(2,1/10+0.011,'10 years', col = 'grey' )
lines(seq(min(log10(data$spVL))-0.5,max(log10(data$spVL))+0.5,0.01), rep(1/15, length(seq(min(log10(data$spVL))-0.5,max(log10(data$spVL))+0.5,0.01) ) ),
      lwd=1 , col = 'grey' )
text(2,1/15+0.011,'15 years', col = 'grey' )
lines(seq(min(log10(data$spVL))-0.5,max(log10(data$spVL))+0.5,0.01), rep(1/5, length(seq(min(log10(data$spVL))-0.5,max(log10(data$spVL))+0.5,0.01) ) ),
      lwd=1, col = 'grey'  )
text(2,1/5+0.011,'5 years', col = 'grey' )
lines(seq(min(log10(data$spVL))-0.5,max(log10(data$spVL))+0.5,0.01), rep(1/2, length(seq(min(log10(data$spVL))-0.5,max(log10(data$spVL))+0.5,0.01) ) ),
      lwd=1, col = 'grey' )
text(2,1/2+0.011,'2 years', col = 'grey')
dev.off()

```




